In-silico molecular enrichment and clearance of the human intracranial space

### 2024-11-27 paper/supplementary_main.tex
Updates from Overleaf

CSF pulsates in the SAS and ventricular system with each cardiac and respiratory cycle. Taking the large spatial differences of pulsatile CSF flow into account, we estimate a local diffusion enhancement factor $R$ in the CSF-filled space $\Omega_{\rm CSF}$ by combining a computational model of cardiac- and respiration-induced pulsatile CSF flow with theoretical estimates for shear-augmented (Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion}.

\rami{I rewrote the above paragraph and some aspects of this section}

\rami{CSF pulsates in the SAS and ventricular system with each cardiac and respiratory cycle with large spatial differences.  This leads to dispersion effects. For example, 
Sharp et al \cite{keith2019dispersion} estimated the enhancement of diffusion in the spinal subarachnoid space due to dispersive effects to average 5.8 times that of purely molecular diffusion with a theoretical maximum of $10^6$. 
This phenomena is accounted for in our model by scaling the free diffusion coeffiecient $D^{\mathrm{Gad}}$ by the formula $D^{\mathrm{eff}} = (1+R) D^{\mathrm{Gad}}$ where $R$ represents a local diffusion enhancement in the CSF-filled space $\Omega_{\rm CSF}$.}

\rami{ Note that since there is no data for the local values of $R$, we estimate $R$ by combining computational results (solving Stokes flow equations, see below) and explicit estimates for dispersion enhancement that are derived under simplifying assumptions on the problem geometry, the flow field and concentration profiles \cite{watson1983diffusion, keith2019dispersion}. Therefore such an estimate is heuristic; however,  our computational results qualitatively agree with observed data.  A key feature in our estimate of $R$ is that it is spatially varying, including local effects depending on a  computed flow field. }

Similar to the case of CSF flow driven by production introduced in \Cref{sec:csf_flow}, we employ a Stokes flow model to predict CSF velocities corresponding to the peak cardiac and respiratory blood inflow rates. In particular, we enforce a uniformly distributed inflow of $u^{\rm in}_{\rm AM}$ and $u^{\rm in}_{\rm LV}$ across the arachnoid mater ($\Gamma_{\rm AM}$) and lateral ventricular surfaces ($\Gamma_{\rm LV}$) respectively, mimicking the effect of reduced CSF volume due to blood inflow. We set $u^{\rm in}_{\rm AM} = 6\,$ml/s \cite{causemann2022human,baledent2014imaging} and $u^{\rm in}_{\rm LV}=0.31\,$ml/s \cite{vinje2019respiratory} in the cardiac case, and $u^{\rm{in}}_{\rm{AM}} = 1\,$ml/s \cite{gutierrez2022effect} and $u^{\rm in}_{\rm{LV}}=0.121\,$ml/s \cite{liu2024using} in the respiratory case, respectively.
We prescribe a no-slip boundary condition on the pial surface $\Gamma_{\rm Pia}$ and zero-traction on the spinal boundary $\Gamma_{\rm SSAS}$, making the spinal compartment the only outflow route of CSF. Thus, we compute the pulsatile flow fields by augmenting the Stokes flow model (eq. \ref{eq:momnetum_equation} and \ref{eq:divergence_equation}) with the following boundary conditions:
\begin{equation}\label{eq:cardiac_flow_bcs}
    \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
    \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
    \bm u = \frac{u^{\rm in}_{\rm LV} \cdot \bm n}{|\Gamma_{\rm LV}|} \text{ on } \Gamma_{\rm LV}; \quad
    \bm u = \frac{u^{\rm in}_{\rm AM} \cdot \bm n}{|\Gamma_{\rm skull}|} \text{ on } \Gamma_{\rm AM}.
\end{equation}
The obtained pressure fields $p_{\rm cardiac}$ and $p_{\rm resp}$ only account for viscous forces, albeit cardiac and respiratory induced CSF flow is known to be unsteady and substantially impacted by transient inertial forces. Assuming an oscillatory flow with either the heartbeat frequency $\omega_{\rm cardiac} = 2 \pi$ or respiratory frequency $\omega_{\rm resp} = 0.5 \pi$, a mean gap width of the CSF-filled spaces of $2h=3$\,mm and CSF density $\rho=10^3\,\text{kg/m}^3$, the squared Womersley numbers $\alpha_i^2 = \frac{h^2 \omega_{i} \rho}{\mu}$ become $\alpha_{\rm cardiac}^2 = 20.2$ and $\alpha_{\rm resp}^2=5.05$, which is unsteady. Based on theoretical considerations on the ratio of oscillatory flow to steady flow impedances in a tube, we account for inertial forces in the pulsatile flow by upscaling the computed viscous pressure field by a factor of $1 + \alpha^2 / 8$ (see \cite{van1998cardiovascular}, chapter. 4.3). Thus, the upscaled pressure fields become $p^{\rm up}_{i}= (1 + \alpha_{i}^2 / 8)p_i \text{ for } i = \{\mathrm{cardiac}, \mathrm{resp}\}$.

>>>>

CSF pulsates in the SAS and ventricular system with each cardiac and respiratory cycle, leading to an increase in effective solute diffusivity via a process known as Taylor dispersion \cite{taylor1953dispersion, watson1983diffusion}. Previous estimates of the magnitude of this effect in the CSF-filled space range from an enhancement factor of 5.8 \cite{keith2019dispersion} up to $10^4$ \cite{ray2021quantitative}. The large discrepancies in reported estimates can (at least partly) be attributed to different assumptions on representative values for local width, pressure differences and fluid velocities in the SAS, which reflect the diverse flow characteristics in the different regions of the intracranial CSF-filled space.

Taking the large spatial differences of pulsatile CSF flow into account, we estimate a spatially varying diffusion enhancement factor $R$ in the CSF-filled space $\Omega_{\rm CSF}$ by combining a computational model of cardiac- and respiration-induced pulsatile CSF flow with theoretical estimates for shear-augmented (Taylor) dispersion \cite{taylor1953dispersion, watson1983diffusion}. These estimates are are derived under simplifying assumptions on the problem geometry, the flow field and concentration profiles \cite{watson1983diffusion, keith2019dispersion}. Therefore such an estimate is heuristic; however, our computational results qualitatively agree with observed data.

Similar to the case of CSF flow driven by production introduced in \Cref{sec:csf_flow}, we employ a Stokes flow model to predict CSF velocities corresponding to the peak cardiac and respiratory blood inflow rates. In particular, we enforce a uniformly distributed inflow of $u^{\rm in}_{\rm AM}$ and $u^{\rm in}_{\rm LV}$ across the arachnoid mater ($\Gamma_{\rm AM}$) and lateral ventricular surfaces ($\Gamma_{\rm LV}$) respectively, mimicking the effect of reduced CSF volume due to blood inflow. We set $u^{\rm in}_{\rm AM} = 6\,$ml/s \cite{causemann2022human,baledent2014imaging} and $u^{\rm in}_{\rm LV}=0.31\,$ml/s \cite{vinje2019respiratory} in the cardiac case, and $u^{\rm{in}}_{\rm{AM}} = 1\,$ml/s \cite{gutierrez2022effect} and $u^{\rm in}_{\rm{LV}}=0.121\,$ml/s \cite{liu2024using} in the respiratory case, respectively.
We prescribe a no-slip boundary condition on the pial surface $\Gamma_{\rm Pia}$ and zero-traction on the spinal boundary $\Gamma_{\rm SSAS}$, making the spinal compartment the only outflow route of CSF. Thus, we compute the pulsatile flow fields by augmenting the Stokes flow model (eq. \ref{eq:momnetum_equation} and \ref{eq:divergence_equation}) with the following boundary conditions:
\begin{equation}\label{eq:cardiac_flow_bcs}
    \bm u = 0 \text{ on } \Gamma_{\rm Pia}; \quad 
    \mu \nabla \bm u \cdot \bm n - p \bm n = 0 \text{ on } \Gamma_{\rm SSAS}; \quad 
    \bm u = \frac{u^{\rm in}_{\rm LV} \cdot \bm n}{|\Gamma_{\rm LV}|} \text{ on } \Gamma_{\rm LV}; \quad
    \bm u = \frac{u^{\rm in}_{\rm AM} \cdot \bm n}{|\Gamma_{\rm skull}|} \text{ on } \Gamma_{\rm AM}.
\end{equation}
The obtained pressure fields $p_{\rm cardiac}$ and $p_{\rm resp}$ only account for viscous forces, albeit cardiac and respiratory induced CSF flow is known to be unsteady and substantially impacted by transient inertial forces. Assuming an oscillatory flow with either the heartbeat frequency $\omega_{\rm cardiac} = 2 \pi$ or respiratory frequency $\omega_{\rm resp} = 0.5 \pi$, a mean gap width of the CSF-filled spaces of $2h=3$\,mm and CSF density $\rho=10^3\,\text{kg/m}^3$, the squared Womersley number $\alpha_i^2 = \frac{h^2 \omega_{i} \rho}{\mu}$ becomes $\alpha_{\rm cardiac}^2 = 20.2$ and $\alpha_{\rm resp}^2=5.05$, which is unsteady. Based on theoretical considerations on the ratio of oscillatory flow to steady flow impedances in a tube, we account for inertial forces in the pulsatile flow by upscaling the computed viscous pressure field by a factor of $1 + \alpha^2 / 8$ (see \cite{van1998cardiovascular}, chapter. 4.3). Thus, the upscaled pressure fields become $p^{\rm up}_{i}= (1 + \alpha_{i}^2 / 8)p_i \text{ for } i = \{\mathrm{cardiac}, \mathrm{resp}\}$.
